Abstract Data Type Inference

### 2015-03-18 matching.tex
A reasonable story for matching scheme computation.

TODO ARGUE THAT SIMPLE NORMALIZING MATCHING SCHEMES CAN BE
COMPUTED/APPROXIMATED BY CHECKING WHETHER FINITE SETS OF EXECUTIONS AND THEIR
PROJECTIONS ARE MATCH-REMOVAL CLOSED

\begin{theorem}

  The matching scheme inference problem is computable for \ldots

\end{theorem}

>>>>

We automate the computation of matching schemes by considering the function
${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
logical satisfiability problem, and constrain ${\sf g}$ to normalize the
executions of an implementation. To this end, we fix an implementation
$\mathcal{I}$ and a language $L = \tup{E,P,Q}$ to which $\mathcal{I}$ adheres,
then consider any enumeration $F$ of execution pairs $\tup{e,e'} \in E^2$ such
that
\begin{itemize}

  \item $e \in \mathcal{I}$ and $e' \not\in \mathcal{I}$, and

  \item $e'$ is obtained by deleting operations of $e$.

\end{itemize}
Any such pair of executions can be used to rule out several possibilities,
for instance, that the operations removed from $e$ to obtain $e'$:
\begin{itemize}

  \item do not constitute a match in $e$,

  \item are not all duplicate operations,

  \item are not all read-only operations,

  \item do not constitute multiple matches in $e$,

\end{itemize}
and so on. Ruling out these possibilities is sound since, for example, a
normalized scheme $M$ could not consider those operations a match: otherwise
the history abstraction $H(\mathcal{I},M)$ which includes $H(e,M)$ must also
include $H(e',M)$, being normal, and in particular closed under match removal.
Such an $M$ would thus not be faithful. For simplicity, in what follows we
consider ruling out only the first possibility: that the operations removed
from $e$ are not a match. In principle, the approach extends to rule out all
possibilities.

In what follows, we denote the operations of an execution $e$ by $O_e$.
To consider whether a given pair $o_1, o_2$ of operations of an execution $e$
is a match according to the simple matching scheme based on ${\sf g}$, for
the given language $L$, we define the macro $\textsc{IsMatch}_{e,o_1,o_2}$:
\begin{align*}
  \mathsf{g}\big( P(e,o_1), P(e,o_2), Q(e,o_1,o_2) \big).
\end{align*}
Then a given set $O \subseteq O_e$ constitutes a match according to
${\sf g}$ when all operations $o_2 \in O$ target some operation $o_1 \in O$,
and no other operation $o_2 \in O \setminus O_e$ does. We express this with the
macro $\textsc{EntireMatch}_{e,O}$:
\begin{align*}
  \bigvee_{o_1 \in O}
  \Big(
  \bigwedge_{o_2 \in O} \textsc{IsMatch}_{e,o_1,o_2}
  \land \bigwedge_{o_2 \in O_e \setminus O} \lnot \textsc{IsMatch}_{e,o_1,o_2}
  \Big)
\end{align*}
Finally, given a pair $\tup{e,e'} \in F$, we prohibit the operations $O_e
\setminus O_{e'}$ from constituting a match according to ${\sf g}$, since if
${\sf g}$ normalized $\mathcal{I}$, and $e \in \mathcal{I}$, then $e'$ should
also be in $\mathcal{I}$. We express this exclusion for all pairs of $F$
with the macro $\textsc{Normalizes}_{F}$:
\begin{align*}
  \bigwedge_{\tup{e,e'}\in F} \lnot \textsc{EntireMatch}_{e,(O_e \setminus O_{e'})}
\end{align*}
We thus check satisfiability for the conjunction of non-matches $O_e \setminus
O_{e'}$ over all pairs $\tup{e,e'} \in F$.

On the one hand, checking satisfiability of $\textsc{Normalizes}_F$ can be used
to conclude the impossibility of a good matching scheme — at least for the
given language.

\begin{lemma}

  If $\textsc{Normalizes}_F$ is unsatisfiable, then there exists no simple
  matching scheme that normalizes $\mathcal{I}$ for the language $L$.

\end{lemma}

In this case, the reason for unsatisfiability can be used as a counterexample
to refine the given language, for example, by adding additional predicates.

  TODO SHOW THAT THE REGISTER LANGUAGE EXAMPLE WITHOUT ONE OF THE PREDICATES
  WILL LEAD TO UNSAT.

On the other hand, the satisfiability of $\textsc{Normalizes}_F$ does not
necessarily lead to a unique matching scheme, since $\textsc{Normalize}_F$ can
have multiple satisfying assignments. Furthermore, $\textsc{Normalizes}_F$ does
not necessarily normalize $\mathcal{I}$. For one reason, $F$ may not be a
complete set of examples of executions and invalid projections. Second, our
simple characterization of $\textsc{Normalize}_F$ does not rule out the other
reasons for a given example $\tup{e,e'} \in F$ to be an invalid projection,
e.g.,~that the removed operations do not constitute \emph{multiple} matches.
However, we believe that checking satisfiability of $\textsc{Normalize}_F$ is
useful nonetheless: at the very least, satisfying assignments can be used as
\emph{assistance} in constructing normalizing matching schemes.
